---
title: "CO2 Prediction Analysis"
author: "Shubhan Tamhane"
date: "2025-01-18"
output: pdf_document
---

```{r setup, include=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org/"))
knitr::opts_chunk$set(echo = TRUE)
```
Including Plots

Importing neccesary libraries 
```{r}
library(dplyr)
library(car)
library(randomForest)
```

Importing data 
```{r}
d1 <- read.csv("/Users/shubhantamhane/Downloads/CO2 Emissions.csv")
colnames(d1)
```
Checking for null values 
```{r}
sum(is.na.data.frame(d1))
```

Exploring data

```{r}
head(d1)
```


```{r}
str(d1)
```

```{r}
unique(d1$Vehicle.Class)
```
Data Preprocessing 

High correlations between features, could cause collinearity issues 
```{r}
cor(d1$Engine.Size.L.,d1$Cylinders)
```

```{r}
cor(d1$Engine.Size.L.,d1$CO2.Emissions.g.km.)
```

```{r}
cor(d1$Cylinders, d1$CO2.Emissions.g.km.)
```

Since I want to avoid multicollinearity and engine size has a higher correlation with the target variable, I chose to drop the cylinders feature. 

```{r}
d1 <- d1 %>% select(-Cylinders)
```


```{r}
unique(d1$Transmission)
```

Since there were too many models and it would be too difficult to encode them, I decided to drop the feature. 
```{r}
d1 <- d1 %>% select(-Model)
```

```{r}
colnames(d1)
```

```{r}
unique(d1$Fuel.Type)
```

Drop all types of gas mileage except for miles per gallon, as this is the industry norm. 
```{r}
d1 <- d1 %>% select(-Fuel.Consumption.City..L.100.km., -Fuel.Consumption.Hwy..L.100.km., -Fuel.Consumption.Comb..L.100.km.)
```

```{r}
colnames(d1)
```


```{r}
unique(d1$Make)
```

Separate all car makes into either economy or luxury brands. A value of 0 is given to all economy cars and a 1 to all luxury brands. Add this column to the d1 dataframe. 
```{r}
economy_brands <- c("BUICK", "CHEVROLET", "CHRYSLER", "DODGE", "FIAT", "FORD", "GMC", "HONDA", "HYUNDAI", "JEEP", "KIA", "MAZDA", "MINI", "MITSUBISHI", "NISSAN", "SCION", "SMART", "SUBARU", "TOYOTA", "VOLKSWAGEN", "RAM")


luxury_brands <- c("ACURA", "ALFA ROMEO", "ASTON MARTIN", "AUDI", "BENTLEY", "BMW", "CADILLAC", "CADILLAC", "JAGUAR", "LAMBORGHINI", "LAND ROVER", "LEXUS", "LINCOLN", "MASERATI", "MERCEDES-BENZ", "PORSCHE", "ROLLS-ROYCE", "SRT", "VOLVO", "GENESIS", "BUGATTI")

d1$brand_encoded <- ifelse(d1$Make %in% economy_brands, 0, 1)

```

```{r}
temp <- tapply(d1$CO2.Emissions.g.km., d1$brand_encoded, mean, na.rm = TRUE)
barplot(temp, 
        main = "CO2 Emissions by Luxury/Economy Brand", 
        xlab = "Type of Make", 
        ylab = "CO2 Emissions (g/km)", 
        las = 2)
```

```{r}
unique(d1$Transmission)
```

To encode the different types of transmissions, I separated them into the following columns: Automatic sequential, automatic, manual, automated manual, and continuously variable transmission. I then encoded these and added them to the d1 dataframe. 

```{r}
Automatic_seq <- c("AS5", "AS6", "AS8", "AS7", "AS9", "AS4", "AS10")
Automatic <- c("A6", "A7", "A8", "A4", "A5", "A9")
Manual <- c("M6", "M7", "M5")
Automated_Manual <- c("AM6", "AM7", "AM5", "AM8", "AM9")

d1$Transmission_encoded <- ifelse(d1$Transmission %in% Automatic_seq, 1, 
                                  ifelse(d1$Transmission %in% Automatic, 2, 
                                         ifelse(d1$Transmission %in% Manual, 3,
                                                ifelse(d1$Transmission %in% Automated_Manual, 4, 0))))
```

```{r}
#Boxplot for vehicle transmission types
boxplot(d1$CO2.Emissions.g.km. ~ d1$Transmission_encoded, 
        main = "CO2 Emissions by Vehicle Transmission", 
        xlab = "Vehicle Transmission Type", 
        ylab = "CO2 Emissions (g/km)", las = 2)
```


```{r}
unique(d1$Vehicle.Class)
```

I separated the cars into coupes, sedans, SUVs, vans and trucks as these are the most common types of cars. I then encoded these and added them to the d1 dataframe. 

```{r}
coupe <- c("TWO-SEATER", "MINICOMPACT", "SUBCOMPACT")
sedan <- c("COMPACT", "MID-SIZE", "FULL-SIZE")
SUV <- c("SUV - SMALL", "SUV - STANDARD")
VAN <- c("VAN - CARGO", "VAN - PASSENGER", "MINIVAN")
Truck <- c("PICKUP TRUCK - SMALL", "PICKUP TRUCK - STANDARD")

d1$Class_encoded <- ifelse(d1$Vehicle.Class %in% coupe, 1, 
                           ifelse(d1$Vehicle.Class %in% sedan, 2, 
                                  ifelse(d1$Vehicle.Class %in% SUV, 3,
                                         ifelse(d1$Vehicle.Class %in% VAN, 4,  
                                                ifelse(d1$Transmission %in% Truck,5, 0)))))

d1 <- subset(d1, Class_encoded != 0)
```

```{r}
# Boxplot to see emissions by Vehicle Class
boxplot(d1$CO2.Emissions.g.km ~ d1$Class_encoded, 
        main = "CO2 Emissions by Vehicle Class", 
        xlab = "Vehicle Class", ylab = "CO2 Emissions (g/km)", las = 2)
```

I changed the fuel type from a string to a factor variable to encode it. The fuel had many types including normal gas, premium gas, no gas (electric), diesel, and ethanol. 
```{r}
d1$Fuel.Type <- as.factor(d1$Fuel.Type)
```

```{r}
# Boxplot to see emissions by Fuel Type
boxplot(d1$CO2.Emissions.g.km ~ d1$Fuel.Type, 
        main = "CO2 Emissions by Fuel Type", 
        xlab = "Fuel Type", ylab = "CO2 Emissions (g/km)", las = 2)
```

Fitting linear model
```{r}
m1 <- lm(CO2.Emissions.g.km. ~ Class_encoded + Transmission_encoded + brand_encoded + Fuel.Consumption.Comb..mpg. +  Fuel.Type, data = d1)
```

```{r}
summary(m1)
```

```{r}
residualPlots(m1, type = "rstudent")
```
Since the graph displays heavy curvature, a linear model does not seem to be appropriate here. 

```{r}
studentized_residuals <- rstudent(m1)
hist(studentized_residuals)
```




```{r}
model_rf <- randomForest(d1$CO2.Emissions.g.km. ~ Class_encoded + Transmission_encoded + brand_encoded + Fuel.Consumption.Comb..mpg. +  Fuel.Type, data = d1, ntree = 500)
summary(model_rf)
```


```{r}
predictions <- predict(model_rf, d1)
r_squared <- cor(d1$CO2.Emissions.g.km, predictions)^2
print(paste("R-Squared:", round(r_squared, 4)))
```
















